Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-05-31 main.tex
Figure updates and text for figures.

In the early years of developments automatic control, \citep{Whipple1899}
remarkably realized and showed that roll motion feedback can stabilize the a
bicycle. Much later, automatic roll stabilization of single track vehicles
began in earnest after predictive motorcycle models were developed and refined
in throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's
work and seems to be the first \todo{Check if a Japanese researcher was first}
to attempt to robotically stabilize a small motorbike with a controlled
inverted pendulum that mimicked rider lean, but he was not successful in
demonstrating what his theoretic control model correctly predicted. In his
model, he recognized that feedback of the vehicle roll angle and angular rate
was essential to stabilize the vehicle. It was not until the mid 1980s that
Ruijs and Pacjecka successfully demonstrated an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
propose how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when one is not
concerned with wobble instabilities, which are not prominent in bicycles at low
speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
stability and their vehicle was fully automatic, no human rider was involved.
Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single actor falls. Most
of these robotic bicycle and motorcycle designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to change the dynamics,
up to some limits, of the human controlled plant. In our prior study, we
demonstrate reduced motion during distractions~\cite{Alizadehsaravi2023} due to
the balance assist but conversely researchers with a similar vehicle recently
showed rider disatisfction with the stabilization~\cite{Hanakam2023}.

@@

\section{Methods}
%
\subsection{Bicycle}
%
We modified a standard electric bicycle (Gazelle Grenoble C7, Dieren, The
Netherlands) with a custom motor in the steering assembly capable of applying
up to 7~\si{\newton\meter} of torque between the head tube and steer tube. A
custom motor controller converts the commanded torque to applied torque. We
measure the speed of the rear wheel with an encoder (Magura XXX) and measure
the body fixed roll rate of the bicycle with a MEMs rate gyroscope (Company,
Location).  The balance assist control algorithm is implemented on a
microprocessor (Teensy, Company, Location) and data from all sensors is logged
with a CAN bus (Company, Location) at approximately X Hz.

@@

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem. The four motors are programmed to apply a light force at all
times to keep the ropes taught and to track a commanded force profile using a
PID controller running on a microprocessor (Arduino XXX, Italy). We measure the
force applied by each motor at the handlebar via four inline load cells, each
with a max load of XX~\si{\newton}. The commanded force profiles are designed
to apply an external pulsive torque to the front assembly (handlebars, forks,
wheel) at magnitudes varying from X to X. The four motors are arranged at the
four corners of a 1~\si{\meter} wide treadmill (make, Company, City) that can
reach speeds of 18~\si{\kilo\meter\per\hour}. The general design is described
in detail in the Van De Velde's MSc thesis~\cite{vandeVelde2022a}. Our
modifications relative to that described in the thesis include simplifying the
controller with a inexpensive microcontroller and the use of a simpler safety
harness.

\begin{figure}
  \centering
  % TODO : scale these pngs to a reasonable size for an A4 sheet
  \subcaptionbox{}{\includegraphics[width=70mm]{figures/participant-in-set-up.png}}
  \subcaptionbox{Diagram of perturbation system.}{\includegraphics[width=70mm]{figures/setup-diagram.png}
}
\end{figure}

\begin{figure}
  \includegraphics[width=\textwidth]{figures/perturbation_0.png}
  \caption{Force nd torque plts of a pterubation of 80N per rope in the
  counterclockwise direction (from Marten's thesis, pg 26)}
\end{figure}

\begin{figure}
  \centering
  \subcaptionbox{Start}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Thesis Fig 5.1 on pg 37}
\end{figure}

@@

\subsection{Protocol}
%
The subjects were divided into two groups. Tthe first group performed the
protocol at a belt speed of 10~\si{\kph} with the controller gain as $k=8$ and
the second group performed the protocol at a belt speed of 6~\si{\kph}. The
10~\si{\kph} group's experiments occured X weeks before the
6~\si{\kilo\meter\per\hour} group.

Subjects wore a helmet and fall safety harness attached to the ceiling. We
allowed them practice riding on the treadmill until they indicated they were
comfortable enough to have perturbations applied. For most, this was less than a
10~\si{\minute} warm up. We then asked the rider to ride for 90~\si{\second},
attempting to maintain the location of their front wheel on the center line of
the treadmill. We define a ``fall'' on the treadmill by two criteria: the rider
removes their foot from the pedal and places it on the ground or the bicycle
wheel exceeds the width of the treadmill. We then applied perturbations in
random directions (clockwise or counter clockwise) increasing the magnitude by
X~\si{\newton\meter} each time. We log the magnitude that causes the first fall
to characterize that subject's nominal fall threshold. Following this we apply
20 perturbations of random magnitudes to the cyclist while they ride at a
constant speed and record which perturbations cause a fall. We let the cyclist
rest and then perform another 20 perturbations. We randomize whether the
balance assist system is on or off during the first or second set of
perturbations.

>>>>

In the early years of developments automatic control, \citep{Whipple1899}
remarkably realized and showed that roll motion feedback can stabilize the a
bicycle. Much later, automatic roll stabilization of single track vehicles
began in earnest after predictive motorcycle models were developed and refined
in throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's
work and seems to be the first to attempt to robotically stabilize a small
motorbike with a controlled inverted pendulum that mimicked rider lean, but he
was not successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that \cite{Nagai1983} sucessfully demonstrated balancing a robotic
bicycle on a treadmill with both steering control and a laterally moving mass.
Ruijs and Pacjecka followed this by demonstrating an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
propose how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when one is not
concerned with wobble instabilities, which are not prominent in bicycles at low
speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
stability and their vehicle was fully automatic, no human rider was involved.
Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single actor falls. Most
of these robotic bicycle and motorcycle designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to change the dynamics,
up to some limits, of the human controlled plant. In our prior study, we
demonstrate reduced motion during distractions~\cite{Alizadehsaravi2023} due to
the balance assist but conversely researchers with a similar vehicle recently
showed rider disatisfction with the stabilization~\cite{Hanakam2023}.

@@

\section{Methods}
%
\subsection{Bicycle}
%
We modified a standard electric bicycle (Gazelle Grenoble C7, Dieren, The
Netherlands) with a custom motor in the steering assembly capable of applying
up to 7~\si{\newton\meter} of torque between the head tube and steer tube,
Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller converts the
commanded torque to applied torque. We measure the speed of the rear wheel with
an encoder (Magura XXX) and measure the body fixed roll rate of the bicycle
with a MEMs rate gyroscope (Company, Location). The balance assist control
algorithm is implemented on a microprocessor (Teensy, Company, Location) and
data from all sensors is logged with a CAN bus (Company, Location) at
approximately X Hz.
%
\begin{figure}
  \centering
  \includegraphics[width=100mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with steer motor and data
    aquisition and control electronics mounted in the rear rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem, Figure~\ref{fig:setup-diagram}. The four motors are
programmed to apply a light force at all times to keep the ropes taught and to
track a commanded force profile using a PID controller running on a
microprocessor (Arduino XXX, Italy). We measure the force applied by each motor
at the handlebar via four inline load cells, each with a max load of
XX~\si{\newton}. The commanded force profiles are designed to apply an external
pulsive torque to the front assembly (handlebars, forks, wheel) at magnitudes
varying from X to X. The four motors are arranged at the four corners of a
1~\si{\meter} wide treadmill (make, Company, City) that can reach speeds of
18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
Van De Velde's MSc thesis~\citep{vandeVelde2022a} and the physical arrangment
is shown in Figure~\ref{fig:participant-in-set-up}. Our modifications relative
to that described in the thesis include simplifying the controller with a
inexpensive microcontroller and the use of a simpler safety harness.
%
\begin{figure}
  \centering
  % TODO : scale these pngs to a reasonable size for an A4 sheet
  \subcaptionbox{
    A participant on the bicycle in the safety harness with the Bump'Em motor
    lines attached to the ends of the handlebars.
    \label{fig:participant-in-set-up}
  }{
    \includegraphics[width=70mm]{figures/participant-in-set-up.png}
  }
  \subcaptionbox{
    Top view diagram of bicycle handlebar perturbation system.
    \label{fig:setup-diagram}
  }{
    \includegraphics[width=70mm]{figures/setup-diagram.png
  }
}
\end{figure}

\begin{figure}
  \includegraphics[width=150mm]{figures/perturbation_0.png}
  \caption{Motor force and resulting  steer torque perturbation based on an 80N
  counterclockwise applied force.}
\end{figure}

@@

\todo[inline]{Include the average age, mass, height and standard deviation if
we recorded those.}

\subsection{Protocol}
%
The subjects were divided into two groups. The first group performed the
protocol at a belt speed of 2.8~\si{\meter\per\second}
(10~\si{\kilo\meter\per\hour}) with the controller gain at $k=8$ and the second
group performed the protocol at a belt speed of 1.7~\si{\meter\per\second}
(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occured X
weeks before the 6~\si{\kilo\meter\per\hour} group.

Subjects wore a helmet and fall safety harness attached to the ceiling
Figure~\ref{fig:participant-in-set-up}. We allowed them to practice riding on
the treadmill until they indicated they were comfortable enough to have
perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
We then asked the rider to ride for 90~\si{\second}, attempting to maintain the
location of their front wheel on the center line of the treadmill. We define a
``fall'' on the treadmill by two criteria: the rider removes their foot from
the pedal and places it on the ground or the bicycle wheel exceeds the width of
the treadmill. Figure~\ref{fig:fall} shows an example fall. We then applied
perturbations in random directions (clockwise or counter clockwise) increasing
the magnitude by X~\si{\newton\meter} each time.  We log the magnitude that
causes the first fall to characterize that subject's nominal fall threshold.
Following this we apply 20 perturbations of random magnitudes to the cyclist
while they ride at a constant speed and record which perturbations cause a
fall. We let the cyclist rest and then perform another 20 perturbations. We
randomize whether the balance assist system is on or off during the first or
second set of perturbations.
%
\begin{figure}
  \centering
  \subcaptionbox{Start}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Thesis Fig 5.1 on pg 37}
  \label{fig:fall}
\end{figure}
